---
title: "An Analysis of Taylor Swift's Lyrics"
author: "Rachael Olson"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: FALSE

library(tidyverse)
library(tidytext)
library(textdata)
library(wordcloud)
library(wordcloud2)
library(igraph)
library(ggraph)
library(dplyr)
library(viridis)
library(ggthemes)
library(stopwords)
```


# Introduction

Taylor Swift is known for her emotional ties to her words, detailed storytelling, empowering songs, and the various genres she has covered throughout her time in the music industry. In this analysis, I will examine her words through frequency, emotions, empowering themes, and other components. This data was acquired from Alice Zhao's github page(adashofdata), where Zhao is  a data scientist and an instructor at Northwestern University. This data only contains music from her first 10 albums and the lyrics are only containing parts of her music. Alice Zhao gathered the data through the Genius Lyrics API and Spotify API. Throughout this analysis, this data will be demonstrated through various regular expressions, text analysis, and string operations. 


```{r}
#load dataset
library(readxl)
ts_data <- read_excel("~/Desktop/SDS 264/Taylor_Swift_Genius_Data.xlsx")

#smart stop words
smart_stop_words <- get_stopwords(source = "smart")

#data cleaning: three strings and two regex
ts_data <- ts_data |>
  mutate(Lyrics = str_to_lower(Lyrics), #string the lyrics to lowercase
         Lyrics = str_replace_all(Lyrics, "\\d+",""), #replace numbers
         Lyrics = str_replace_all(Lyrics, "\\b\\w+'\\w+\\b", " ")) #removes words with apostrophes
```


# First Plot: Analysis of Commonly Used Word by Taylor Swift
```{r}
#| message: FALSE
#initial analysis of commonly used words by Taylor Swift
ts_data |>
  unnest_tokens(word, Lyrics) |> #splits lyrics into tokens
  anti_join(smart_stop_words) |> #eliminates commonly used words
  count(word, sort = TRUE) |> #counts and sorts from most to least used words
  slice_max(n, n = 20) |> 
  ggplot(aes(x= fct_reorder(word, n), y = n, fill  = n)) + #shows a commonly used words top to bottom
  geom_col() +
  coord_flip() +
  labs(title =  "Most Frequent Words in Taylor Swift's Songs",
       x = "Words",
       y = "Frequency",
       fill = "Frequency",
       caption = "Data from: https://github.com/adashofdata/taylor_swift_data.git") +
  scale_fill_gradient(low = "#f9c5d1", high = "#d50073") +  # Light pink to hot pink
  theme_classic()

```

This plot demonstrates that words such as *love* and *time* are her most frequently used words despite commonly used words such as I and my. This demonstrates emotional connection in her words as these two suggest both affection and reflection.


# Second Plot: Positive and Negatively Related Words in Taylor Swift's Songs
```{r}
#| message: FALSE
#| warning: FALSE

bing_sentiments <- get_sentiments(lexicon = "bing") #binary(+/-)
nrc_sentiments <- get_sentiments(lexicon = "nrc") #emotion words

#sentiment analysis one
ts_data |>
  unnest_tokens(word, Lyrics) |> #splits lyrics into tokens
  anti_join(smart_stop_words) |> #get rid of the stop words
  inner_join(bing_sentiments) |> #+/-
  count(sentiment, word, sort = TRUE) |>
  group_by(sentiment) |>
  slice_max(n, n = 10) |>
  ungroup() |>
  ggplot(aes(x = fct_reorder(word, n), y = n, fill = sentiment)) +
    geom_col() +  
    coord_flip() +
    facet_wrap(~ sentiment, scales = "free") +
  labs(title = "Positively and Negatively Used Words in Taylor Swift's Songs",
       x = "Words",
       y = "Frequency",
       fill = "Word Sentiment",
       caption = "Data from: https://github.com/adashofdata/taylor_swift_data.git") +
  theme_bw() +
  scale_fill_manual(values = c(positive = "#009E60", negative = "#D70040"))
```

# Third Plot: Emotions Expressed in Taylor Swift's Lyrics
```{r}
#| message: FALSE
#| warning: FALSE

#sentiment analysis two
ts_data |>
  unnest_tokens(word, Lyrics) |>
  inner_join(nrc_sentiments) |>
  count(sentiment, sort = TRUE) |>
  ggplot(aes(x = fct_reorder(sentiment, n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  coord_flip () +
  labs(title = "Expressed Emotions in Taylor Swift's Lyrics",
       x = "Emotions",
       y = "Frequency",
       caption = "Data from: https://github.com/adashofdata/taylor_swift_data.git") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw()
```

These past two graphs demonstrates Taylor Swift's lyrical blend that contains joyful notes with emotional notes. This represents her duality in sentiment that contributes to why so many people relate to her music.

# Word Cloud: Empowering Words in Taylor Swift's Lyrics
```{r}
#| message: FALSE
# regex use
# only filtered for commonly associated words to empowerment
empower_words <- ts_data |>
  unnest_tokens(word, Lyrics) |> #words from lyrics
  filter(str_detect(word,"\\b(fight|strong|stand|rise|power|worthy|enough|brave|smart|work|confident|strength|advocate|courageous)\\b")) |> #filter for empowerment words
  count(word, sort = TRUE)

#text analysis: wordcloud
wordcloud(
  words = empower_words$word, 
  freq = empower_words$n, 
  scale = c(4.5, 1),
  colors = brewer.pal(6, "Accent"),
  max.words = 10, 
  random.order = TRUE)
```

In this word cloud, words such as enough, fight, brave, and strong are just a few strongly empowering words that are represented throughout her music. These words are not solely representing the grasp of the lyrics themselves, as words are typically more powerful together, as well as give a clearer picture of the words they are associated with. 

# Fourth Plot: Two Word Links Within Taylor Swift's Music
```{r}
#| message: FALSE

ts_two <- ts_data |> #making every two words its own row
  unnest_tokens(bigram, Lyrics, token = "ngrams", n = 2) |> #two words together
  filter(bigram != "NA") |>
  separate(bigram, c("word_one", "word_two"), sep = " ") |> #seperate two words
  filter(!word_one %in% smart_stop_words$word,
         !word_two %in% smart_stop_words$word) |> #filter for stop words
  count(word_one, word_two, sort = TRUE)

#lots of repeated words back to back
ts_two |> head(20)

#two word graph
two_graph <- ts_two |>
  filter(word_one != word_two) |>
  filter(n > 8) |>
  graph_from_data_frame()

#directed tree of two word links
ggraph(two_graph, layout = "fr") +
  geom_edge_link(show.legend = FALSE, color = "hotpink")+ #lines; no legend
  geom_node_point(color = "pink", size = 5) + #point markers
  geom_node_text(aes(label = name), vjust = 1.5, hjust = 1.2, color = "darkgreen") + #word labels
  theme_void()
```

This graph represents words that are correlated together more than eight times. Eliminating stop words and repeated words helps demonstrate the more commonly correlated words together. Demonstrating poetic phrases and repeated phrases gives more insight into the lyrics and how they are created and placed together to make a Taylor Swift song. 

# Conclusion

This analysis looked through various strings, regular expressions, and text analysis to examine Taylor Swift's crafting of her words within her first ten albums. From romantic themes to emotional lows to empowering chants, there is a wide range of words and themes of words throughout her career. The most frequently used words, sentiments, and phrases all are powerful examples of the strength it takes to change genres over the various years while still maintaining a common theme throughout her storytelling. Storytelling is the common theme when it comes to Taylor Swift's writing as there are various emotions throughout each album and lyric.

## References

Adashofdata. (n.d.). Adashofdata/taylor_swift_data: Song and Lyric Data collected for Taylor Swift’s 10 studio albums (up through Midnights). GitHub. https://github.com/adashofdata/taylor_swift_data.git 
